The Effects of Racial Identity and Discrimination on Political Participation and Political Attitudes

### 2024-12-16 Final Report/ForSubmission_IDS702_finalreport.qmd
cleaning files

---
title: "The Effects of Racial Identity and Discrimination on Political Participation and Political Attitudes"
author: "Peter de Guzman, Jennifer Li, Nakiyah Dhariwala, Sizhe Chen"
format: pdf
date: 2024-12-08
editor: visual
---

@@

```{r echo=FALSE, results=FALSE}

@@

# Add row names as a new column
analysis_table <- cbind(RowName = names, results_base_model)

# Render the table
table <- analysis_table |>
  gt() |>
  tab_header(
    title = "Odds Ratios with Confidence Intervals and P-Values for base model"
  ) |>
  cols_label(
    RowName = "Description",  # Change the column header for the row names
    OddsRatio = "Odds Ratio",
    LowerCI = "Lower Confidence Interval",
    UpperCI = "Upper Confidence Interval",
    PValue = "P-Value"
  ) |>
  fmt_number(
    columns = c(OddsRatio, LowerCI, UpperCI),
    decimals = 4
  ) |>
  fmt_scientific(
    columns = c(PValue),
    decimals = 2  # Show p-values in scientific notation
  )

# View the table
#table

# Update the column names in the dataframe
colnames(analysis_table) <- c("Description", "Odds Ratio", "Lower Confidence Interval", 
                              "Upper Confidence Interval", "P-Value")

# View analysis table
#analysis_table

# View the updated dataframe
#head(analysis_table)

@@

Following the Immigration and Nationality Act of 1965, which allowed for increased immigration into the United States, the United States has become increasingly racially and ethnically diverse. However, studies of political participation and public opinion have often failed to include large samples of racial and ethnic minority Americans, limiting researchers from drawing conclusions on the attitudes of these groups. Using survey data from the 2020 Collaborative Multi-Racial Post-Election Survey, which includes oversamples of racial and ethnic minority populations in the U.S., we investigate the impact of racial identity and racism on political participation and public opinion. First, we analyze whether valuing belonging in a racial group can influence a respondent’s interest in politics. Then, we analyze how personal experiences with racism may influence how respondents value policies aimed at addressing racial and ethnic discrimination. Our findings suggest that racial group belonging is not associated with higher odds  of increased political interest, while gender identity, racial identity, and party affiliation may be associated with higher levels of political interest. Additionally, gender identity, racial identity, and personal experiences with discrimination may be associated with higher odds of increased levels of support for anti-discrimination policies.

@@

Our analysis focuses on two topics within the field of racial and ethnic politics.

@@

To investigate these two research questions, we used survey data from the 2020 Collaborative Multi-Racial Post-Election Survey (CMPS) \[7\]. The CMPS is an extensive dataset designed to capture perspectives across racial and ethnic groups in the United States. The survey was offered in multiple languages, including English, Spanish, Chinese, Korean, Vietnamese, Arabic, Urdu, Farsi, and Haitian Creole.

@@

We streamlined the dataset by consolidating religious affiliations into a single categorical variable, Religion, with a new category, Multiple Religions, for respondents reporting more than one affiliation. Variables with over 50% missing values were excluded. Key demographic variables, including educational attainment, sexual orientation, and rurality were re-coded into simplified categories. Age and race were structured into meaningful groups. Extraneous variables were removed, and categorical variables were converted to ordered factors to ensure consistency and clarity for analysis.

To model the outcome variable, interest in participating in politics, in the first research question, we will use an ordinal logistic regression with predictors including the importance of belonging to a racial group to one’s identity, race, age, educational attainment, and political party affiliation. The outcome variable has levels “Not at all interested in politics”, “Not that interested in politics”, “Somewhat interested”, and “Very interested in politics”. We will also include an interaction term between gender and the importance of belonging to a racial group to one’s identity to assess whether the effect of the importance of belonging to a racial group on one’s identity differs based on one’s gender identity. To assess the model we will use a confusion matrix to compare predicted and observed values, and compare the model’s accuracy and no information rate. Finally, we will conduct a Brant test to provide additional evidence regarding whether the proportional odds assumption was violated.

For the second research question, we will make use of a binomial logistic regression to model the outcome variable ’Importance placed on anti-discrimination policies’ with predictors including race, gender identity, sexual orientation, age, education, community type, experiences with racism, and perception of racism for different groups. Additionally, we will include an interaction term between gender and race to explore intersectionality theory and determine which model performs better. After selecting the better model, we will evaluate and assess its performance using various tests such as the confusion matrix, ROC, AUC, accuracy, and other relevant metrics.

@@

### Research Question 1:

@@

| Response to Q29                   | Count  | Share of Total Respondents |
|-----------------------------------|--------|----------------------------|
| Very interested in politics       | 4,306  | 25%                        |
| Somewhat interested               | 7,612  | 43%                        |
| Not that interested in politics   | 3,643  | 21%                        |
| Not at all interested in politics | 1,984  | 11%                        |
| Total                             | 17,545 | 100%                       |

@@

The model also included multiple predictor variables that control for the impact of other aspects of a respondent’s identity, socioeconomic status, and political preferences These included demographic variables such as the primary race reported by a respondent, their age, and gender, as well as a respondent’s reported level of educational attainment and political party affiliation. Finally, the model included an interaction term between gender and the importance a respondent placed in belonging to their racial group to investigate whether the relationship between racial belonging and political interest differs depending on gender. This interaction term allows us to evaluate if the effect of racial belonging on political interest is different for males, females, non-binary respondents, or respondents who specified a different gender identity.

@@

  # Create a mapping for variable names with grouping prefixes
  rename_map <- c(
    # Education Group
    "educCollege graduate" = "1. Education: College Graduate",
    "educSome college experience" = "1. Education: Some College",

    # Gender Group
    "S3B(2) Woman" = "2. Gender: Woman",
    "S3B(3) Non-binary" = "2. Gender: Non-binary",

    # Age Group
    "age_recoded70+" = "3. Age: 70+",
    "age_recoded60-69" = "3. Age: 60-69",
    "age_recoded50-59" = "3. Age: 50-59",

    # Race Group
    "race_recodedBlack" = "4. Race: Black",
    "race_recodedAsian American" = "4. Race: Asian",
    "race_recodedLatino" = "4. Race: Latino",
    "race_recodedAIAN" = "4. Race: Native",
    "race_recodedNHPI" = "4. Race: NHPI",

    # Racial Belonging Group
    "Q271_reordered(2) Very important" = "5. Racial Belonging: Very Important",
    "Q271_reordered(1) Extremely important" = "5. Racial Belonging: Extremely Important",

    # Party ID Group
    "Q21(2) Democrat" = "6. Party ID: Democrat",
    "Q21(3) Independent" = "6. Party ID: Independent",
    "Q21(4) Other party" = "6. Party ID: Other",

    # Interactions
    "Q271_reordered(1) Extremely important:S3B(2) Woman" = "7. Interaction: Racial Belonging & Gender: Woman"
  )

@@

  # Custom sorting function
  sort_by_group_and_name <- function(x) {
    groups <- as.numeric(sub("^(\\d+)\\..+$", "\\1", x))
    names <- sub("^\\d+\\.[^:]+: (.+)$", "\\1", x)
    sort_df <- data.frame(
      original = x,
      group = groups,
      name = names,
      stringsAsFactors = FALSE
    )
    sort_df <- sort_df[order(sort_df$group, -xtfrm(sort_df$name)), ]
    return(sort_df$original)
  }

@@

  # Create plot
  ggplot(plot_data, aes(y = name)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 0.7) +
    geom_point(aes(x = or), size = 4, color = "blue") +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.3, color = "black") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 6, color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.title.x = element_text(size = 12, face = "bold"),
      axis.title.y = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.margin = margin(1, 1, 1, 1, "cm"),
      panel.grid.major.y = element_line(color = "gray90"),
      panel.grid.minor.y = element_blank()
    ) +
    labs(
      title = "Significant Variables with Confidence Intervals",
      x = "Odds Ratio",
      y = "Variables"
    )
}

@@

Note: Some variable levels are excluded due to filtering for significant variables. To view the reference level for each variable group, please see the below table.

| Variable Group | Reference Level |
|------------------------------------|------------------------------------|
| 1\. Education | No college experience |
| 2\. Gender | Man |
| 3\. Age | 18-29 years old |
| 4\. Race | White |
| 5\. Racial Belonging | Not At All Important |
| 6\. Party ID | Republican |
| 7\. Interaction: Racial Belonging & Gender: Woman | Man, Not At All Important |

@@

INSERT TABLE

The values for the sensitivity and specificity of the model for each class are present in the above table. The sensitivity of the model was low (below 0.20) for the levels “Very interested in politics”, “Not that interested in politics”, and “Not at all interested in politics.”. The only level with a high sensitivity value was the response option of being “Somewhat interested” in politics, which had an associated sensitivity of 0.90. These results indicate that the model struggled to correctly identify respondents who were “Very interested”, “Not that interested”, or “Not at all interested” in politics. The specificity of the model was high for each level, except for the response option of being “Somewhat interested” in politics (Specificity: 0.16), indicating that the model classified a high rate of false positives for this level, and it was potentially overly biased to predicting that respondents were “Somewhat interested” in politics. We also produced a confusion matrix for a multinomial model with the same predictor variables. The confusion matrices were very similar, providing evidence that there is not a sizable difference in performance between the ordinal model and the multinomial model. In addition, we conducted a Brant test. The Brant test result indicated that there is not strong evidence that the proportional odds assumption was violated.

### Research Question 2:

@@

We assessed multicollinearity in both models by calculating the normalized Variance Inflation Factor (VIF) values. In all cases, the VIF values (APPENDIX TABLE have made a vif at the end of the R code) were consistently below 2, indicating minimal correlation among predictors, suggesting that both models are robust and not significantly affected by multicollinearity.

To further assess the goodness-of-fit for the two models, a chi-square test was conducted with their residual degrees of freedom and deviance for were as follows: - Base Model: Residual Degrees of Freedom = 17,487, Residual Deviance = 17,934 - Interaction Model: Residual Degrees of Freedom = 17,469, Residual Deviance = 17,913

The difference in deviance between the two models was 21.145 with 18 degrees of freedom, indicating the interaction model to be slightly superior. However, the p-value of 0.2722 exceeded the 0.05 significance level, suggesting that including the interaction term did not significantly improve the model fit. Consequently, the base model was selected as it was more parsimonious with comparable explanatory power.

The findings from the base model revealed that personal experiences with discrimination and certain demographic factors significantly influence the importance respondents place on policies aimed at addressing racial and ethnic discrimination.

```{r, echo=FALSE}

# Filter for significant variables (p-value < 0.05)
significant_data <- analysis_table[analysis_table$`P-Value` < 0.05, ]

@@

# Create the plot
ggplot(significant_data, aes(x = `Odds Ratio`, y = Variable)) +
  geom_point(size = 4, color = "blue") +  # Add points for odds ratios
  geom_errorbarh(aes(xmin = `Lower Confidence Interval`, 
                     xmax = `Upper Confidence Interval`), 
                 height = 0.3, color = "black") +  # Add horizontal error bars
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 0.7) +  # Add a vertical reference line at Odds Ratio = 1
  labs(
    title = "Odds Ratio of Significant Variables",
    x = "Odds Ratio",
    y = "Variable"
  ) +
  theme_minimal() +  # Use a minimal theme for better aesthetics
  theme(
    axis.text.y = element_text(size = 6, color = "black"),  # Improve y-axis text readability
    axis.title.x = element_text(size = 12, face = "bold"),  # Highlight x-axis title
    axis.title.y = element_text(size = 12, face = "bold"),  # Highlight y-axis title
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Center the title
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Adjust margins
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Reduce x-axis text size
    legend.text = element_text(size = 8),  # Reduce legend text size
    legend.title = element_text(size = 9)  # Reduce legend title size
  )

# Save the plot with larger dimensions
#ggsave("Significant_Variables_Plot_Enhanced.png", width = 40, height = 20)

@@

As shown in the figure above, gender and race were the most consistent and influential predictors. Women had 1.25 times higher odds of prioritizing anti-discrimination policies than men, while Black respondents had 3.53 times higher odds than White respondents. Latino and Asian American respondents had 1.74 and 2.15 times higher odds, respectively, to emphasize such policies compared to their White counterparts. These results underscore the significance of addressing the unique experiences of marginalized groups, whose direct experiences with racism may shape their policy attitudes.

Additionally, individuals who perceive higher levels of discrimination against racial or ethnic groups tend to prioritize anti-discrimination policies. For example, those who believed there was "a little" discrimination against Blacks were 45% less likely to prioritize these policies compared to those who saw "a lot" of discrimination, demonstrating that personal perceptions of racial injustice are key in shaping views on anti-discrimination measures. In contrast, education and sexual orientation did not have a statistically significant impact on prioritizing anti-discrimination policies. This reflects that, in this model, personal experiences of discrimination and perceptions of racial inequality have a more direct effect on attitudes toward such policies. While education and sexual orientation may influence political views in other contexts, they seem less relevant to attitudes on racial discrimination. The lack of significance for these variables may indicate that perceptions of racial or ethnic discrimination are stronger motivators for supporting anti-discrimination policies than other demographic factors.

The base model's evaluation (as shown in Table MODEL) provides insights into how personal experiences with discrimination and demographic factors influence the prioritization of anti-discrimination policies. The model achieved an overall accuracy of 74.48%, which closely aligns with the No Information Rate (NIR) of 74.25%, indicating that its performance is not significantly better than predicting the majority class.

From the confusion matrix in Table CONFUSION, it is evident that the model performs well in identifying respondents who did not prioritize anti-discrimination policies ("Not Important"), as reflected by its high specificity of 94.89%., However, it struggled with the sensitivity (recall) for the "Important" class, which was only 15.63%. This low sensitivity indicates that the model missed a large proportion of respondents who do prioritize these policies. Precision for the "Important" class was moderate at 51.46%, but the F1 score of 0.2397 highlights poor overall performance in identifying the minority class. The model's AUC (Area Under the Curve) value (see Fig. 2) of 0.72 suggests that it has a moderate ability to discriminate between the two classes, though it still falls short of ideal performance.

The balanced accuracy was 55.26%, reflecting the model’s challenges in handling class imbalance. Despite achieving statistical significance with McNemar’s test (p-value \< 2e-16), the model’s predictions deviate considerably from the true data. In conclusion, while the model effectively identifies respondents who do not prioritize anti-discrimination policies, it performs poorly in predicting those who do. This imbalance signals the need for improvement in predicting the minority class.

@@

In this analysis, we investigated whether valuing belonging in a racial group can influence a respondent’s interest in politics. We found that the higher levels of importance an individual placed in belonging to their specific racial group (reporting that belonging to their group was “Very important” or “Extremely important”) had a significant positive effect on their interest in politics when compared to male respondents who did not place a high level of importance on belonging to their specific racial group. Further research is needed to understand how an individual’s racial belonging can impact their political participation while controlling for other demographic characteristics.

Following the resource model of participation, age and increases in educational attainment appeared to be associated with higher odds of reporting a higher level of interest in politics. The most significant positive predictors of increased levels of interest in politics were an individual’s age and level of educational attainment. In particular, individuals who graduated from college had much higher odds of reporting a higher level of interest in politics than individuals with no college experience. This follows popular findings that educational attainment can lead to increases in political participation \[8\], and that older Americans vote at higher rates than Americans ages 18-29 years old \[9\].

@@

Despite the value of this survey’s oversamples of racial and ethnic minority communities, some limitations were present in our study. For the ordinal logistic regression, we used educational attainment as a proxy for socioeconomic status. This is a commonly used proxy measure in behavioral and social sciences \[10\]. Further research on the influence of socioeconomic status of political participation for racial/ethnic minorities and the use of other measures such as household income or wealth would help us understand if the aforementioned resource model of participation is indeed generalizable to other groups in the American electorate. The effect of race on political interest appeared to be more pronounced for members of very small racial/ethnic minority groups in our analysis such as Asian Americans, Native Americans, Native Hawaiians and Pacific Islanders. Both models had low specificity, suggesting that they struggled to correctly identify cases where respondents had low levels of political interest or respondents who considered anti-discrimination policies to be important. Similarly, in our second research question, while the model effectively identified respondents who do not prioritize anti-discrimination policies, it struggled to predict those who do. This imbalance underscores the challenges of accurately modeling the minority class—respondents who view anti-discrimination policies as important—and highlights the need for more nuanced approaches to capture these perspectives.

Overall, these challenges and opportunities highlight the importance of disaggregated data and further research in partnership with small racial and ethnic minority communities within the United States. Future work on political participation and public opinion among racial and ethnic minorities could support researchers, policymakers, and advocates in designing more inclusive policies and interventions. Additionally, incorporating qualitative methodologies, such as interviews or community-based participatory research, could provide richer context and amplify the voices of underrepresented communities. Longitudinal studies could also complement these efforts, capturing changes over time and offering deeper insights into the lived experiences of marginalized communities. By addressing these gaps, future research can better capture the diverse and intersectional realities of minority groups and contribute to more effective advocacy and policy efforts.

@@

### Ordinal Model Overall Statistics

|                           | Metric           |
|---------------------------|------------------|
| Accuracy                  | 0.4503           |
| 95% CI                    | (0.4429, 0.4577) |
| No Information Rate (NIR) | 0.4339           |
| P Value \[Acc \> NIR\]    | 6.105e-06        |
| Kappa                     | 0.0787           |
| Mcnemar's Test P-Value    | \<2.2e-16        |

### Confusion Matrix

>>>>

---
title: "The Effects of Racial Identity and Discrimination on Political Participation and Political Attitudes"
author: "Peter de Guzman, Jennifer Li, Nakiyah Dhariwala, Sizhe Chen"
format: pdf
date: 2024-12-14
editor: visual
---

@@

```{r, echo=FALSE, results=FALSE}

@@

# Add row names as a new column
analysis_table <- cbind(RowName = names, results_base_model)
colnames(analysis_table) <- c("Description", "Odds Ratio", "Lower Confidence Interval", 
                              "Upper Confidence Interval", "P-Value")

# View the updated dataframe
#head(analysis_table)
```

@@

Following the Immigration and Nationality Act of 1965, which allowed for increased immigration into the United States, the United States has become increasingly racially and ethnically diverse. However, studies of political participation and public opinion have often failed to include large samples of racial and ethnic minority Americans, limiting researchers from drawing conclusions on the attitudes of these groups. Using survey data from the 2020 Collaborative Multi-Racial Post-Election Survey, which includes oversamples of racial and ethnic minority populations in the U.S., we investigate the impact of racial identity and racism on political participation and public opinion. First, we analyze whether valuing belonging in a racial group can influence a respondent’s interest in politics. Then, we analyze how personal experiences with racism may influence how respondents value policies aimed at addressing racial and ethnic discrimination. Our findings suggest that racial group belonging may be associated with higher odds of increased political interest, while gender identity, racial identity, and party affiliation may be associated with higher levels of political interest. Additionally, gender identity, racial identity, and personal experiences with discrimination may be associated with higher odds of increased levels of support for anti-discrimination policies.

@@

Our analysis focuses on two questions within the field of racial and ethnic politics.

@@

To investigate these two research questions, we used survey data from the 2020 Collaborative Multi-Racial Post-Election Survey (CMPS) \[7\]. The CMPS is an extensive dataset designed to capture perspectives across racial and ethnic groups in the United States offered in multiple languages, including English, Spanish, Chinese, Korean, Vietnamese, Arabic, Urdu, Farsi, and Haitian Creole.

@@

To prepare the data, variables with over 50% missing values were excluded. Key demographic variables, including educational attainment, sexual orientation, and rurality were re-coded into simplified categories. Age and race were structured into meaningful groups. Extraneous variables were removed, and categorical variables were converted to ordered factors to ensure consistency and clarity for analysis.

To model the outcome variable, interest in participating in politics, in the first research question, we will use an ordinal logistic regression with predictors including the importance of belonging to a racial group to one’s identity, race, age, educational attainment, and political party affiliation. The outcome variable has 4 ordered levels: “Not at all interested in politics”, “Not that interested in politics”, “Somewhat interested”, and “Very interested in politics”. We will also include an interaction term between gender and the importance of belonging to a racial group to one’s identity to assess whether the effect of the importance of belonging to a racial group on one’s identity differs based on one’s gender identity. To assess the model we will use a confusion matrix to compare predicted and observed values, and compare the model’s accuracy and no information rate. Finally, we will conduct a Brant test to provide additional evidence regarding whether the proportional odds assumption was violated.

For the second research question, which examines the importance placed on anti-discrimination policies, we will use a binomial logistic regression. The predictors include race, gender identity, sexual orientation, age, education, community type (i.e. rurality), experiences with racism, and perceptions of racism affecting different groups. To further explore intersectionality theory, we will develop a second model that includes an interaction term between gender and race. After identifying the better-performing model, we will evaluate its performance using a range of metrics, including the confusion matrix, accuracy, and other relevant indicators.

@@

@@

**Table 1.1: Breakdown of respondents' interest levels in politics**

|          Response to Q29          | Count  | Share of Total Respondents |
|:---------------------------------:|:------:|:--------------------------:|
|    Very interested in politics    | 4,306  |            25%             |
|        Somewhat interested        | 7,612  |            43%             |
|  Not that interested in politics  | 3,643  |            21%             |
| Not at all interested in politics | 1,984  |            11%             |
|               Total               | 17,545 |            100%            |

@@

The model included multiple predictor variables that control for the impact of other aspects of a respondent’s identity, socioeconomic status, and political preferences. These included demographics such as the primary race reported by a respondent, their age, and gender, as well as a respondent’s reported level of educational attainment and political party affiliation. Finally, the model included an interaction term between gender and the importance a respondent placed in belonging to their racial group to investigate whether the relationship between racial belonging and political interest differs depending on gender. This interaction term allows us to evaluate if the effect of racial belonging on political interest is different for males, females, non-binary respondents, or respondents who specified a different gender identity.

@@

  # Create a mapping for variable names with grouping prefixes
  rename_map <- c(
    # Education Group
    "educCollege graduate" = "Education: College Graduate",
    "educSome college experience" = "Education: Some College",

    # Gender Group
    "S3B(2) Woman" = "Gender: Woman",
    "S3B(3) Non-binary" = "Gender: Non-binary",

    # Age Group
    "age_recoded70+" = "Age: 70+",
    "age_recoded60-69" = "Age: 60-69",
    "age_recoded50-59" = "Age: 50-59",

    # Race Group
    "race_recodedBlack" = "Race: Black",
    "race_recodedAsian American" = "Race: Asian",
    "race_recodedLatino" = "Race: Latino",
    "race_recodedAIAN" = "Race: Native",
    "race_recodedNHPI" = "Race: NHPI",

    # Racial Belonging Group
    "Q271_reordered(2) Very important" = "Racial Belonging: Very Important",
    "Q271_reordered(1) Extremely important" = "Racial Belonging: Extremely Important",

    # Party ID Group
    "Q21(2) Democrat" = "Party ID: Democrat",
    "Q21(3) Independent" = "Party ID: Independent",
    "Q21(4) Other party" = "Party ID: Other",

    # Interactions
    "Q271_reordered(1) Extremely important:S3B(2) Woman" = "Interaction: Racial Belonging & Gender: Woman"
  )

@@

  # Custom sorting function that sorts by the group name only
  sort_by_group_and_name <- function(x) {
    group_name <- sub("^\\d+\\. (.*): .*$", "\\1", x)
    variable_name <- sub("^\\d+\\. .+: (.*)$", "\\1", x)

    # Create a data frame to help with sorting
    sort_df <- data.frame(
      original = x,
      group = group_name,
      name = variable_name,
      stringsAsFactors = FALSE
    )

    # Sort first by group, then by variable name
    sort_df <- sort_df[order(sort_df$group, sort_df$name), ]
    return(sort_df$original)
  }

@@

  # Create plot
  ggplot(plot_data, aes(y = name)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 0.7) +
    geom_point(aes(x = or), size = 3, color = "blue") +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.3, color = "black") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 6, color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.caption = element_text(size = 9, face = "italic", hjust = 0.5),  # Center the caption
      plot.margin = margin(1, 1, 1, 1, "cm"),
      panel.grid.major.y = element_line(color = "gray90"),
      panel.grid.minor.y = element_blank()
    ) +
    labs(
      title = "Odds Ratios (Significant Variables)",
      x = "Odds Ratio",
      y = "Variables",
      caption = "Fig. 1.1: Odds ratio for variables with significant associations"
    )
}

@@

*Note: Some variable levels are excluded due to filtering for significant variables. To view the reference level for each variable group, please see table A1 in the Appendix.*

@@

The values for the sensitivity and specificity of the model for each class are present in table A4 in the Appendix. The sensitivity of the model was low (below 0.20) for the levels “Very interested in politics”, “Not that interested in politics”, and “Not at all interested in politics.” The only level with a high sensitivity value was the response option of being “Somewhat interested” in politics, which had an associated sensitivity of 0.90. These results indicate that the model struggled to correctly identify respondents who were “Very interested”, “Not that interested”, or “Not at all interested” in politics. The specificity of the model was high for each level, except for the response option of being “Somewhat interested” in politics (Specificity: 0.16), indicating that the model classified a high rate of false positives for this level, and it was potentially overly biased to predicting that respondents were “Somewhat interested” in politics. We also produced a confusion matrix for a multinomial model with the same predictor variables. The confusion matrices were very similar, providing evidence that there is not a sizable difference in performance between the ordinal model and the multinomial model. In addition, we conducted a Brant test. The Brant test result indicated that there is not strong evidence that the proportional odds assumption was violated.

@@

To evaluate multicollinearity in both models, we calculated the normalized Variance Inflation Factor (VIF) values. As detailed in Appendix Table B1, all normalized VIF values were consistently below 2, indicating robustness of both models with minimal correlation among predictors, thereby ensuring the reliability of the estimated coefficients.

To further assess the goodness-of-fit for the two models, a chi-square test was conducted with their residual degrees of freedom and deviance for were as follows:

-   **Base Model:** Residual Degrees of Freedom = 17,487, Residual Deviance = 17,934

-   **Interaction Model:** Residual Degrees of Freedom = 17,469, Residual Deviance = 17,913

The difference in deviance between the two models was 21.145 with 18 degrees of freedom, indicating the interaction model to be slightly superior. However, the p-value of 0.2722 exceeded the 0.05 significance level, suggesting that including the interaction term did not significantly improve the model fit. Consequently, the base model was selected as it was more parsimonious with comparable explanatory power. The findings from the base model revealed that personal experiences with discrimination and certain demographic factors significantly influence the importance respondents place on policies aimed at addressing racial and ethnic discrimination.

```{r, echo=FALSE}
# Filter for significant variables (p-value < 0.05)
significant_data <- analysis_table[analysis_table$`P-Value` < 0.05, ]

@@

# Create the plot
ggplot(significant_data, aes(x = `Odds Ratio`, y = Variable)) +
  geom_point(size = 2.5, color = "blue") +  # Add points for odds ratios
  geom_errorbarh(aes(xmin = `Lower Confidence Interval`, 
                     xmax = `Upper Confidence Interval`), 
                 height = 0.3, color = "black") +  # Add horizontal error bars
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 0.7) +  # Add a vertical reference line at Odds Ratio = 1
  labs(
    title = "Odds Ratios (Significant Variables)",
    x = "Odds Ratio",
    y = "Variable",
    caption = "Fig. 2.1: Odds ratio for variables with significant associations"
  ) +
  theme_minimal() +  # Minimal theme for clarity
  theme(
    axis.text.y = element_text(size = 6, color = "black", hjust = 1),  # Adjust y-axis text size and alignment
    axis.title.x = element_text(size = 12),  # X-axis title
    axis.title.y = element_text(size = 12),  # Y-axis title
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),  # Center title
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Adjust margins
    plot.caption = element_text(size = 7, face = "italic", hjust = 0.5),  # Center caption
    axis.text.x = element_text(size = 8)  # X-axis text size
  ) +
  coord_cartesian(clip = "off")  # Ensure content stays within plot limits

@@

As shown in Fig. 2.1, the variables with significant associations are illustrated. For reference levels of each variable group, please see Table B3 in the Appendix. Overall, gender and race emerged as the most consistent and influential predictors of prioritizing anti-discrimination policies. Women had 1.25 times higher odds of prioritizing anti-discrimination policies than men (CI: 1.16-1.35), while Black respondents had 3.53 times higher odds than White respondents (CI: 3.13-4.00). Latino and Asian American respondents had 1.74 (CI: 1.53-1.98) and 2.15 (CI: 1.89-2.45) times higher odds, respectively, to emphasize such policies compared to their White counterparts.

Additionally, individuals who perceived higher levels of discrimination against racial or ethnic groups tended to prioritize anti-discrimination policies. For example, those who believed there was "a little" discrimination against Blacks were 45% less likely to prioritize these policies compared to those who saw "a lot" of discrimination (CI: 0.47-0.65). In contrast, education and sexual orientation did not have a statistically significant impact on prioritizing anti-discrimination policies. To further assess the base model, several performance metrics summarized in Appendix Table B4 and the confusion matrix below provide valuable insights into its strengths and limitations.

```{r, echo=FALSE, results=FALSE}
# Generating a prediction as binary values (0 or 1), where:
## 1 indicates a predicted probability greater than 0.5 (positive class, "Important").
## 0 indicates a predicted probability less than or equal to 0.5 (negative class, "Not important").
model_prediction <- ifelse(predict(base_model, 
                                   newdata = dataset_clean, 
                                   type = "response") > 0.5, 1, 0)

# Converting the predictions factors
## 0 Not important, 1 Important
model_prediction_factor <- factor(model_prediction, 
                                  levels = c(0, 1), 
                                  labels = c("Not important", "Important"))  # Adjusting labels

# Convert the response variable to a factor with matching levels
response_factor <- factor(dataset_clean$ImportanceOfStoppingDiscriminationAgainstMinorities,
                          levels = c("Not important", "Important"))  # Ensuring matching order

# Checking if factors are identical (you can uncomment this to check the levels)
# levels(model_prediction_factor)
# levels(response_factor)

# Generate confusion matrix (if needed)
confmatrix_base <- confusionMatrix(table(model_prediction_factor, response_factor),
                positive = "Important",
                mode = "everything")
```

**Table 2.1: Confusion Matrix**

|                        |                   |               |
|:----------------------:|:-----------------:|:-------------:|
| **Predicted / Actual** | **Not Important** | **Important** |
|   **Not Important**    |       12361       |     3812      |
|     **Important**      |        666        |      706      |

Overall, the model performed well in identifying respondents who did not prioritize anti-discrimination policies ("Not Important"), as evidenced by its high specificity of 94.89%. However, the sensitivity for the "Important" class was notably low at 15.63%, indicating that the model struggled to accurately classify a substantial proportion of respondents who prioritize these policies. This imbalance underscores a significant limitation in identifying the minority class. While the precision for the "Important" class was moderate at 51.46%, the F1 score of 0.2397 highlighted the model's poor overall performance in addressing this class effectively.

The Area Under the Curve (AUC) value of 0.72 (see Fig. 2.2 below) indicates a moderate capacity to discriminate between the two classes, though it fell short of an optimal level. Additionally, the balanced accuracy of 55.26% reflects the model's challenges in addressing the effects of class imbalance.

```{r, echo=FALSE}
# Generate predicted probabilities (not just binary predictions) from the model
model_probabilities <- predict(base_model, newdata = dataset_clean, type = "response")

# Calculate ROC curve
roc_curve <- roc(response_factor, model_probabilities)

# Calculate AUC (Area Under the Curve)
auc_value <- auc(roc_curve)

# Adjust plot margins to make room for the caption below
par(mar = c(5, 4, 4, 5))  # Keep the right margin moderate

# Plot ROC curve
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)
legend("bottomright", legend = paste("AUC =", round(auc_value, 3)), col = "blue", lwd = 2)

# Add the figure label outside the plot below
mtext("Fig. 2.2: ROC Curve", side = 1, line = 4, cex = 0.9, font = 3, adj = 0)  # Adjust the line parameter
```

Although the model achieved statistical significance as indicated by McNemar’s test in Appendix Table B4 (p-value \< 2e-16), its predictive performance remains suboptimal, with predictions deviating considerably from observed data. While the model demonstrates strong performance in identifying respondents who do not prioritize anti-discrimination policies, its ability to accurately predict those who do prioritize these policies is markedly limited.

@@

In this analysis, we investigated whether valuing belonging in a racial group can influence a respondent’s interest in politics. We found that the higher levels of importance an individual placed in belonging to their specific racial group (reporting that belonging to their group was “Very important” or “Extremely important”) had a significant positive effect on their interest in politics when compared to male respondents who did not place a high level of importance on belonging to their specific racial group. Further research is needed to understand how an individual’s racial belonging can impact their political participation. Following the resource model of participation, age and increases in educational attainment appeared to be associated with higher odds of reporting a higher level of interest in politics. Individuals who graduated from college had much higher odds of reporting a higher level of interest in politics than individuals with no college experience. This follows findings that educational attainment can lead to increases in political participation \[8\], and that older Americans vote at higher rates than younger Americans \[9\].

@@

Despite the value of this survey’s oversamples of racial minorities, some limitations were present in our study. For the first analysis, we used educational attainment as a proxy for socioeconomic status. This is a common practice in social sciences \[10\]. Further research on the influence of socioeconomic status on political participation for racial minorities would help us understand if the resource model of participation is generalizable to other groups in the U.S. Both models had low specificity, suggesting that they struggled to correctly identify cases where respondents had low levels of political interest or respondents who considered anti-discrimination policies to be important. Similarly, in our second research question, while the model effectively identified respondents who do not prioritize anti-discrimination policies, it struggled to predict those who do. This imbalance underscores the challenges of accurately modeling the minority class—respondents who view anti-discrimination policies as important—and highlights the need for more nuanced approaches to capture these perspectives.

Overall, these challenges and opportunities highlight the importance of disaggregated data and further research in partnership with small racial and ethnic minority communities. Future work on political participation and public opinion among racial and ethnic minorities could support researchers, policymakers, and advocates in designing more inclusive policies and interventions. By addressing these gaps, future research can better capture the diverse and intersectional realities of minority groups and contribute to more effective advocacy and policy efforts.

@@

**Table A1: Reference levels for variables**

|              **Variable Group**               |    **Reference Level**    |
|:----------------------------------:|:----------------------------------:|
|                   Education                   |   No college experience   |
|                    Gender                     |            Man            |
|                      Age                      |      18-29 years old      |
|                     Race                      |           White           |
|               Racial Belonging                |   Not At All Important    |
|                   Party ID                    |        Republican         |
| Interaction: Racial Belonging & Gender: Woman | Man, Not At All Important |

**Table A2: Ordinal Model Statistics**

|                           |    **Metric**    |
|:-------------------------:|:----------------:|
|         Accuracy          |      0.4503      |
|          95% CI           | (0.4429, 0.4577) |
| No Information Rate (NIR) |      0.4339      |
|  P Value \[Acc \> NIR\]   |    6.105e-06     |
|           Kappa           |      0.0787      |
|  Mcnemar's Test P-Value   |    \<2.2e-16     |

**Table A3: Confusion Matrix**
